Finch: Sparse and Structured Array Programming with Control Flow

### 2024-04-06 sections/related.tex
fix figures

\paragraph{ Compilers for Structured Data:}
Several compilers exist for several types of structured data, often featuring separate languages for the storage of the structured data and the computation.
%
The TACO compiler originally supported just plain einsum computations~\cite{kjolstad_tensor_2017}, but has been extended several times to support (single dimensional) local tensors \cite{kjolstad_tensor_2019}, imperfectly nested loops \cite{dias_sparselnr_2022}, breaks via semi-rings~\cite{henry_compilation_2021}, windowing and tiling \cite{senanayake2020sparse}, and convolution~\cite{won2023unified}, all as separate extensions.
%
Similarly, TACO originally support just dense and CSF like N dimensional structures, but was extended independently to support COO like structures~\cite{chou2018format}, and tree like structures~\cite{chou2022compilation}, as separate extensions. SparseTIR is a similar system supporting combined 
sparse formats (including block structures) \cite{ye_sparsetir_2022}.
%
The SDQL language offers a similar level of control flow~\cite{shaikhha2022functional}, but only on sparse hash tables.
%
Similarly, SDQL has been extended with a system that allows one to specify formats as queries on a set of base storage types~\cite{schleich2023optimizing} and separately by another system that describes static symmetries and other structures as predicates~\cite{ghorbani2023compiling}.
%
%Taichi is another language with a separate program and data structure specification, but
The Taichi language focuses on a single sparse data structure made from dense blocks, bit-masks, and pointers~\cite{hu_taichi_2019}.
%
%Outside of these works, the window of comparison fogs up.
%
The sparse polyhedral framework builds on CHiLL for the purpose of generating inspector/executor optimizations~\cite{strout2018sparse} though the branch of this work that specifies sparse formats separately from the computation (otherwise they are inlined into the computation manually) seems to apply mainly to einsums~\cite{zhao2022polyhedral}.
%
Second to last, SQL's classical physical/logical distinction is the classic program/format distinction, and SQL supports a huge variety of control flow constructs~\cite{kotlyar1997relational, date1989guide}.
%
However, many SQL or dataframe systems rely on BTrees, columnar, or hash tables, with only a few systems, such as Vectorwise~\cite{boncz2012vectorwise}, LaraDB~\cite{hutchison2017laradb}, GMAP~\cite{tsatalos1996gmap}, or SciDB~\cite{stonebraker2013scidb} building physical layouts with other constructs based in array programming.
%
However, array based databases are a new focus given the rise of mixed ML/DB pipelines~\cite{baumann2021array,luo2018scalable}.
%
Lastly, SPIRAL focuses on recursively defined datastructures and recursively define linear algebra, and can therefore express a structure and computation that none of the systems mentioned above can: a Cooley–Tukey FFT ~\cite{franchetti2018spiral,franchetti2009operator}.
%
%RECUMA AND UNISPARSE are in OOPSLA R1, we should probably address these

>>>>

\paragraph{ Compilers for Structured Data:}
Several compilers exist for several types of structured data, often featuring separate languages for the storage of the structured data and the computation.
%
The TACO compiler originally supported just plain einsum computations~\cite{kjolstad_tensor_2017}, but has been extended several times to support (single dimensional) local tensors \cite{kjolstad_tensor_2019}, imperfectly nested loops \cite{dias_sparselnr_2022}, breaks via semi-rings~\cite{henry_compilation_2021}, windowing and tiling \cite{senanayake2020sparse}, and convolution~\cite{won2023unified}, and compilation in MLIR \cite{bik_compiler_2022}, all as separate extensions.
%
Similarly, TACO originally support just dense and CSF like N dimensional structures, but was extended independently to support COO like structures~\cite{chou2018format}, and tree like structures~\cite{chou2022compilation}, as separate extensions. SparseTIR is a similar system supporting combined 
sparse formats (including block structures) \cite{ye_sparsetir_2022}.
%
The SDQL language offers a similar level of control flow~\cite{shaikhha2022functional}, but only on sparse hash tables.
%
Similarly, SDQL has been extended with a system that allows one to specify formats as queries on a set of base storage types~\cite{schleich2023optimizing} and separately by another system that describes static symmetries and other structures as predicates~\cite{ghorbani2023compiling}.
%
%Taichi is another language with a separate program and data structure specification, but
The Taichi language focuses on a single sparse data structure made from dense blocks, bit-masks, and pointers~\cite{hu_taichi_2019}.
%
%Outside of these works, the window of comparison fogs up.
%
The sparse polyhedral framework builds on CHiLL for the purpose of generating inspector/executor optimizations~\cite{strout2018sparse} though the branch of this work that specifies sparse formats separately from the computation (otherwise they are inlined into the computation manually) seems to apply mainly to einsums~\cite{zhao2022polyhedral}.
%
Second to last, SQL's classical physical/logical distinction is the classic program/format distinction, and SQL supports a huge variety of control flow constructs~\cite{kotlyar1997relational, date1989guide}.
%
However, many SQL or dataframe systems rely on BTrees, columnar, or hash tables, with only a few systems, such as Vectorwise~\cite{boncz2012vectorwise}, LaraDB~\cite{hutchison2017laradb}, GMAP~\cite{tsatalos1996gmap}, or SciDB~\cite{stonebraker2013scidb} building physical layouts with other constructs based in array programming.
%
However, array based databases are a new focus given the rise of mixed ML/DB pipelines~\cite{baumann2021array,luo2018scalable}.
%
Lastly, SPIRAL focuses on recursively defined datastructures and recursively define linear algebra, and can therefore express a structure and computation that none of the systems mentioned above can: a Cooley–Tukey FFT ~\cite{franchetti2018spiral,franchetti2009operator}.
%
%RECUMA AND UNISPARSE are in OOPSLA R1, we should probably address these
